---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ivs

<!-- badges: start -->

[![Codecov test coverage](https://codecov.io/gh/DavisVaughan/ivs/branch/main/graph/badge.svg)](https://app.codecov.io/gh/DavisVaughan/ivs?branch=main) [![R-CMD-check](https://github.com/DavisVaughan/ivs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/DavisVaughan/ivs/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

ivs (said, "eye-vees") is a package dedicated to working with intervals in a generic way.
It introduces a new type, the **i**nterval **v**ector, which is generally referred to as an **iv**.
An iv is created from two parallel vectors representing the starts (inclusive) and ends (exclusive) of the intervals, like this:

```{r}
library(ivs)

# Interval vector of integers
iv(1:5, 7:11)

# Interval vector of dates
starts <- as.Date("2019-01-01") + 0:2
ends <- starts + c(2, 5, 10)

iv(starts, ends)
```

There are a number of useful things you can do with these, including:

-   Determining how two ivs are related (i.e. does one precede, follow, or overlap the other?) with `iv_locate_overlaps()`.

-   Grouping / Merging overlapping intervals within a single iv with `iv_groups()`.

-   Splitting an iv on its overlapping endpoints with `iv_splits()`.

-   Applying set theoretical operations on two ivs, such as `iv_set_intersect()`.

Interval vectors are completely generic, meaning that you can create them from any comparable type that is supported by [vctrs](https://vctrs.r-lib.org).
This means that user defined S3 types work automatically, like `hms::hms()`, `bit64::integer64()`, and `clock::year_month_day()`.

## Learning

The best way to learn about ivs is by reading the [Getting Started](https://davisvaughan.github.io/ivs/articles/ivs.html) vignette!

## Installation

Install the released version from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ivs")
```

You can install the development version of ivs from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("DavisVaughan/ivs")
```

## Inspiration

This package was inspired by many sources!

-   [IRanges](https://github.com/Bioconductor/IRanges) is the closest equivalent, and inspired many of the function names seen here.
    It is mainly focused on integer intervals, and always uses closed intervals.
    It is also based on S4, and unfortunately that currently means it can't be used as a column in a tibble with the current limitations in vctrs.

-   [intervals](https://github.com/edzer/intervals) is another R package that supports intervals.
    It supports integer/numeric intervals and allows for varying the endpoint bounds.

-   [data.table](https://github.com/Rdatatable/data.table) contains a function named `foverlaps()` for detecting overlaps (which was inspired by `IRanges::findOverlaps()`).

-   [Maintaining Knowledge about Temporal Intervals](https://cse.unl.edu/~choueiry/Documents/Allen-CACM1983.pdf) is a paper by James Allen that `iv_locate_relates()` is based on.
    It is a great primer on interval algebra.

-   [Why numbering should start at 0](https://www.cs.utexas.edu/users/EWD/transcriptions/EWD08xx/EWD831.html) is a small white paper that describes why right-open intervals are often the best choice.
